---
title: "Interval estimates of QTL location"
teaching: 0
exercises: 0
questions:
- "?"
objectives:
- 
- 
keypoints:
- "."
- "."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

For the blood pressure phenotype, we’ve seen good evidence for QTL on chromosomes 1 and 4. Interval estimates of the location of QTL are commonly obtained via 1.5-LOD support intervals, which may be calculated via the function `lodint`. Alternatively, an approximate Bayes credible interval may be obtained with `bayesint`.

```{r message=FALSE, echo=FALSE}
hyper <- calc.genoprob(hyper, step=1, error.prob=0.01)
out.hk <- scanone(hyper, method="hk")
```

To obtain the 1.5-LOD support interval and 95% Bayes interval for the QTL on chromosome 4, type:

```{r}
lodint(out.hk, chr=4)
bayesint(out.hk, chr=4)
```

The first and last rows define the ends of the intervals; the middle row is the estimated QTL location.

It is sometimes useful to identify the closest flanking markers; use `expandtomarkers=TRUE`:

```{r}
lodint(out.hk, chr=4, expandtomarkers=TRUE)
bayesint(out.hk, chr=4, expandtomarkers=TRUE)
```

We can calculate the 2-LOD support interval and the 99% Bayes interval as follows.

```{r}
lodint(out.hk, chr=4, drop=2)
bayesint(out.hk, chr=4, prob=0.99)
```

The intervals for the chr 1 locus may be calculated as follows.

```{r}
lodint(out.hk, chr=1)
bayesint(out.hk, chr=1)
```