---
title: "QTL Effects"
teaching: 0
exercises: 0
questions:
- "?"
objectives:
- 
- 
keypoints:
- "."
- "."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
```


We may obtain plots indicating the estimated effects of the QTL via `plotPXG`, which creates a dot plot, or
`effectplot`, which plots the average phenotype for each genotype group.For `plotPXG`, we must first identify the marker closest to the QTL peak. Use
`find.marker`.

```{r}
max(out.hk)
mar <- find.marker(, chr=7, pos=47.7)
plotPXG(hyper, marker=mar)
```

Note that red dots correspond to inferred genotypes (based on a single imputation).

The function `effectplot` uses the multiple imputation results from `sim.geno`.

```{r}
effectplot(hyper, mname1=mar)
```

We may use `effectplot` at a position on the “grid” between markers, using "7@47.7" to indicate the position at 47.7 cM on chr 7.

```{r}
effectplot(hyper, mname1="7@47.7")
```

Similar plots may be obtained for the locus on chr 15.

```{r}
max(out.hk, chr=15)
mar2 <- find.marker(hyper, chr=15, pos=12)
plotPXG(hyper, marker=mar2)
effectplot(hyper, mname1="15@12")
```

We may plot the joint effects of the two loci via
`plotPXG` as follows:

```{r}
plotPXG(hyper, marker=c(mar, mar2))
plotPXG(hyper, marker=c(mar2, mar))
```

The function `effectplot` gives more readable figures in this case; it’s often useful to look at it in both ways.

```{r}
effectplot(hyper, mname1="7@47.7", mname2="15@12")
effectplot(hyper, mname2="7@47.7", mname1="15@12")
```

The two loci do not appear to interact.