---
title: "Performing a genome scan with binary traits"
teaching: 20
exercises: 10
questions:
- "How do I create a genome scan for binary traits?"
objectives:
- Convert phenotypes to binary values.
- Use logistic regression for genome scans with binary traits.
- Plot and compare genome scans for binary traits.
keypoints:
- "A genome scan for binary traits (0 and 1) requires special handling; scans for non-binary traits assume normal variation of the residuals."
- "A genome scan for binary traits  is performed with logistic regression."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("10-")
```

The genome scans above were performed assuming that the residual variation followed a normal distribution. This will often provide reasonable results even if the residuals are not normal, but an important special case is that of a binary trait, with values 0 and 1, which is best treated differently. The `scan1` function can perform a genome scan with binary traits by logistic regression, using the argument `model="binary"`. (The default value for the `model` argument is `"normal"`.) At present, we _can not_ account for relationships among individuals in this analysis.

Let's first turn our two phenotypes into binary traits by thresholding at the median. One would generally _not_ do this in practice; this is just for illustration.

```{r threshold_phenotypes, eval=FALSE}
bin_pheno <- apply(iron$pheno, 2, function(a) as.numeric(a > median(a)))
rownames(bin_pheno) <- rownames(iron$pheno)
```

We now perform the genome scan as before, including `model="binary"` to indicates that the phenotypes are binary traits with values 0 and 1.

```{r binary_trait_scan, eval=FALSE}
out_bin <- scan1(pr, bin_pheno, Xcovar=Xcovar, model="binary")
```

Here is a plot of the two LOD curves.

```{r plot_bin_scan, eval=FALSE}
par(mar=c(5.1, 4.1, 1.1, 1.1))
ymx <- maxlod(out_bin)
plot(out_bin, map, lodcolumn=1, col="slateblue", ylim=c(0, ymx*1.02))
plot(out_bin, map, lodcolumn=2, col="violetred", add=TRUE)
legend("topleft", lwd=2, col=c("slateblue", "violetred"), colnames(out_bin), bg="gray90")
```

We can use `find_peaks` as before.

```{r find_peaks_bin_scan, eval=FALSE}
find_peaks(out_bin, map, threshold=3.5, drop=1.5)
```
